Experience Report: Haskell in Computational Biology

### 2012-03-09 mrfy_experience_report.tex
section 5 done, but it sucks.

\subsection{Our group then}
 - reimplementing algorithms
  - we reimplemented and yet it was easy and pleasant
  - past reimplementation of a variant of Viterbi DP, in haste, -> 700 line C function
  - building on that reimplementation
 - new implementations (Matt)
  - Modifying Formatt took nearly a year
  - mutable data structures hard to repurpose
  - one enhancement deemed infeasible
 - C++ failures most commonly segfaults, memory errors, heap exhaustion

\subsection{Our group now}
 - Higher order functions militate towards more flexibility
 - Haskell failures are usually typecheck failures; at worst runtime errors around bounds checking
 - MRFy about 3 months of part-time work

\subsection{Computational biology at large}
 - some good software, like HMMER

\section{What we think about it}

>>>>

\subsection{Our group then}
Programmers generally avoid reimplementing algorithms, but there may be good reasons for doing so.
Although the HMMER software contains an mature C++ implementation of the Viterbi algorithm,
a previous member of our group implemented \textit{multidimensional} dynamic programming for SMURF
in haste, resulting in 12,000 lines of undocumented code (including one function of 700 lines).
Though we are using the same Viterbi algorithm HMMER implements, we chose to reimplement it because
we wished to explore the design space of stochastic search techniques using higher order functions
in Haskell.
We found that reimplementing the Viterbi algorithm in Haskell was easy and pleasant.
In contrast, the programming effort to transform SMURF into SMURFLite, a minor evolution, required
several months of research to simply understand the SMURF codebase.

A set of enhancements to another large C++ implementation, the Matt~\cite{Menke:2008wu} structural
aligner, required nearly a year of programming due to complex mutable data structures being 
hard to repurpose.
In fact, we deemed one enhancement infeasible.
Our experience working with these C++ codebases largely involved the failure modes of
segmentation faults, memory errors, and heap exhaustion.

\subsection{Our group now}
Implementing MRFy in Haskell allowed us great flexibility to explore the space of stochastic search
techniques, initial guess heuristics, and search termination.
Higher order functions militate towards more flexibility, so that adding several planned features 
to MRFy should be easy.
The first functioning version of MRFy took roughly three months of part-time work.
The Haskell failure modes we have experienced have largely been typecheck errors; at the worst,
we have seen runtime errors around bounds checking of arrays and vectors, though we wanted
better debugging tools for these runtime errors.

\subsection{Computational biology at large}
There exists some excellent software in the computational biology community, though little of it
has been written in Haskell.
The training component of MRFy was modified from that of HMMER, and working with the HMMER
codebase was relatively pleasant, as data structures and their mutators are well documented.
BioHaskell exists (and we use \texttt{Bio.Sequence.Fasta} as a parser) but is not nearly as
complete as BioPython or BioRuby, which are heavily used in the community.
We hope that BioHaskell and other tools for computational biology in Haskell continue to mature.
\section{What we think about it}
